Foreign direct investment, corporate social responsibility, and malaria control in Mozambique - trends, risks, and opportunities

### 2017-12-27 paper.tex
updated

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
%\makeatletter
%\def\ps@pprintTitle{%
% \let\@oddhead\@empty
% \let\@evenhead\@empty
% \def\@oddfoot{\it \hfill\today}%
% \let\@evenfoot\@oddfoot}
%\makeatother

% A modified page layout
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Foreign direct investment, corporate social responsibility, and malaria control in Mozambique - trends, risks, and opportunities},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
\usepackage{longtable}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header
\usepackage{longtable}

\usepackage[nomarkers]{endfloat}

\begin{document}
\begin{frontmatter}

  \title{Foreign direct investment, corporate social responsibility, and malaria
control in Mozambique - trends, risks, and opportunities}
    \author[isglobal,cism,vu]{Joe Brew\corref{c1}}
   \ead{joebrew@economicsofmalaria.com} 
   \cortext[c1]{Corresponding Author}
    \author[isglobal,icl,cism]{Elisa Sicuri}
   \ead{elisa.sicuri@isglobal.org}

      \address[isglobal]{Barcelona Institute for Global Health, c/ Rosselló, 132, 5è 2a. 08036,
Barcelona, Spain}
    \address[icl]{Imperial College London, South Kensington Campus, London SW7 2AZ, U.K.}
    \address[cism]{Centro de Investigação em Saúde de Manhiça, Vila da Manhiça, Bairro
Cambeve, Rua 12, Distrito da Manhiça, CP 1929, Maputo, Mozambique}
    \address[vu]{VU University Amsterdam, De Boelelaan 1105, 1081 HV Amsterdam,
Netherlands}

  \begin{abstract}
  Foreign direct investment (FDI) in Mozambique has increased rapidly in
  the last two decades. The growing interest in corporate social
  responsibility (CSR) - combined with a recent push for malaria
  elimination - suggest the need for a critical examination of the role of
  FDI in malaria control and elimination. Through a systematic review of
  the literature, we find that there has been a notable increase in
  research pertaining to FDI, CSR and malaria in recent years. Consensus
  among researchers is that this increase has not been accompanied by a
  substantial private sector investment in malaria control (with a few
  notable exceptions). This suggests a potential opportunity for scaling
  up privately driven malaria control as a means to both improve public
  health and increase return on investment. However, given the lack of
  coordination between public and private sectors, even when interests
  coincide, an over-reliance on foreign and private initiative for funding
  eradication is not without risks.
  \end{abstract}

 \end{frontmatter}

@@

Mozabique's recent economic growth has been facilitated by the
government's open policy towards foreign investment, a plentiful supply
of natural resources (particularly minerals and hydrocarbons) (Rogers
2014) and relatively inexpensive labor. Improvements in health have
accompanied economic expansion, but Mozambique still lags behind in
basic health outcomes - particularly those related to malaria - even by
regional standards. Most economic activities have to deal with malaria:
particularly agricultural-based and extraction activities cannot be
carried out without consistent investment towards malaria management,
both prevention and prompt treatment. This can be done as part of the
``normal activities'' of private sector or can go slightly beyond what
is strictly needed for the activities and be seen as part of the
corporate social responsibilityDespite strong evidence showing that
improved health yields substantial economic benefits (Brundtland 1999,
Bloom and Canning (2008)), there is little evidence regarding the extent
to which ``corporate social responsibility'' (CSR) activities of foreign
firms have targeted or impacted malaria. Understanding the relationship
between FDI, CSR and malaria is essential to identifying which
incentives and barriers to private sector involvement in malaria
elimination exist, as well as identifying potential opportunities and
synergies for scaling up private sector involvment in malaria
elimination.

That said, The few malaria-related CSR projects in recent years has come
in large, ``mega-projects'' whose well-publicized malaria abatement
activities are profit-driven (Mouzin and al. 2011) or whose primary aim
is not profit-related (Han 2015). Though the former is often portrayed
as a ``win-win'' for business and public health, the latter also offers
tangible benefits for private industry, and should be understood as
operating under the same conditions and with the same motivations.

Private foreign investors showed a great optimism during the FMI meeting
held in Maputo in 2014 (reference needed). Beginning in late 2015, the
flood of FDI into Mozambique (and many other subsaharan African nations)
slowed to a trickle. The impact of this slowdown on CSR is not yet
known, but it can reasonably be assumed that it will mean a reduction in
CSR activities (albeit with lag). Almost in parallel to this, Mozambique
has had serious financial problems which have resulted in default
(reference needed
http://www.economist.com/news/middleeast-and-africa/21715030-mozambique-fails-pay-its-debtsmozambiques-default).
Given the rapidly changing economic and epidemiologic context in
Mozambique, a comprehensive and current understanding of both (a) the
landscape of FDI and CSR in Mozambique and (b) public health issues
(namely, malaria control) which are directly affected those investments
is sorely needed. Such an understanding may foster greater private
interest in for-profit investment in public health measures. Likewise,
it can facilitate better public sector understanding of potential
industry partners and stakeholders (a prerequisite to greater
public-private collaborations), and guide the public sector away from an
inflexible dependence on FDI for the provision of public health
necessities.

@@

We examined and analyzed trends from multiple datasets related to FDI,
CSR and malaria in Mozambique. These included the ``Doing Business''
data pertaining to the measurement of regulatory efficiency, the GADM,
the Deutsche Bundesbank Data Repository for foreign exchange rate
history, the Knoema World Data Atlas for macroeconomic trends, the World
Bank data portal for information containing to net foreign inflows and
FDI, USAID Demographic and Health Survey data on sociodemographics and
health-related practices, the Instituto Nacional de Estatistica for
granular data pertaining to Mozambique's population and economic
activities, and Institute for Health Metrics and Evaluation data
pertaining to disease trends over time.

@@

@@

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-6-1} \end{center}

The massive increases in FDI have also been facilitated by dramatic
decreases in the costs to import and export, as well as the costs of
starting a business and registering property (Bank 2016).

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-7-1} \end{center}

@@

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-8-1} \end{center}

@@

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-9-1} \end{center}

@@

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-10-1} \end{center}

\subsubsection{CSR in Mozambique}\label{csr-in-mozambique}

Firms with CSR activities are often large and foreign. Among the largest
``key players'' in Mozambican CSR are Coca-Cola, British Petroleum and
Colgate-Palmolive. State conglomerates, such as Águas de Moçambique and
Electricidade de Moçambique, also participate in CSR activities (Compact
2007).

\subsubsection{Malaria-related CSR and other corporate
activity}\label{malaria-related-csr-and-other-corporate-activity}

Most CSR-funded activities are focused on education, community
development, women's rights and entrepreneurship. According to a UN
poll, none of the country's largest firms invest directly in
malaria-related CSR activities (Compact 2007). This may be due to the
perceived costs of malaria control, lack of perceived PR benefits, the
government and NGO's predominant role in the area, as well as the
``opt-in'' nature and privacy/legal issues generally related to engaging
in health-related campaigns.

That said, a majority of the firms interviewed by the UN indicated that
one of the principal reasons for investment in CSR is to complement
government efforts. To the extent that malaria accounts for more of the
loss in disability-adjusted life years in Mozambique than comparable
countries (IHME 2015), malaria control's lack of representation among
core CSR activities is a notable absence.

Whether under the guise of CSR or not, it is noteworthy that the private
sector currently does play a role in malaria control activities.
According to 2011 DHS data, greater than 7 in every 1,000 households had
a private company carry out indoor residual spraying (DHS 2011). And, in
some clusters, the percent of houses covered by private IRS was greater
than 25\%.

@@

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-12-1} \end{center}

@@

One notable exception is the Nando's-lead Goodbye Malaria Trust, an
umbrella organization which inclues a development impact bond, the
Goodbye Malaria initiative, and partnerships with several other foreign
firms operating in Mozambique (more details here).

@@

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-15-1} \end{center}

@@

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

@@

@@

\hypertarget{ref-doingbusiness}{}
Bank, World. 2016. \emph{Doing Business 2016: Measuring Regulatory
Quality and Efficiency}. World Bank.
\url{http://www.doingbusiness.org/~/media/GIAWB/Doing\%20Business/Documents/Annual-Reports/English/DB16-Chapters/DB16-Mini-Book.pdf}.

@@

@@

@@

@@

@@

@@

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Foreign direct investment, corporate social responsibility, and malaria control in Mozambique - trends, risks, and opportunities},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Foreign direct investment, corporate social responsibility, and malaria
control in Mozambique - trends, risks, and opportunities}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{longtable}
\usepackage{multicol}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\fancyhead[C]{joebrew@economicsofmalaria.com $\bullet$ ROUGH DRAFT $\bullet$ FDI in Mozambique}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\newcommand{\footremember}[2]{%
    \footnote{#2}
    \newcounter{#1}
    \setcounter{#1}{\value{footnote}}%
}
\newcommand{\footrecall}[1]{%
    \footnotemark[\value{#1}]%
}

\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\widowpenalties 1 150

\makeatletter
\renewcommand\footnotesize{%
   \@setfontsize\footnotesize\@ixpt{11}%
   \abovedisplayskip 8\p@ \@plus2\p@ \@minus4\p@
   \abovedisplayshortskip \z@ \@plus\p@
   \belowdisplayshortskip 4\p@ \@plus2\p@ \@minus2\p@
   \def\@listi{\leftmargin\leftmargini
               \topsep 4\p@ \@plus2\p@ \@minus2\p@
               \parsep 2\p@ \@plus\p@ \@minus\p@
               \itemsep \parsep}%
   \belowdisplayskip \abovedisplayskip
}
\makeatother

\begin{document}
\maketitle

\begin{center}
\begin{large}

Joe Brew\footremember{isglobal}{Barcelona Institute for Global Health: c/ Rosselló, 132, 5è 2a. 08036, Barcelona, Spain}\footremember{cism}{Centro de Investigação em Saúde de Manhiça: Vila da Manhiça, Bairro Cambeve, Rua 12, Distrito da Manhiça, CP 1929, Maputo, Mozambique}\footremember{vu}{VU University Amsterdam: De Boelelaan 1105, 1081 HV Amsterdam, Netherlands}
Celine Aerts\footrecall{isglobal}
Elisa Sicuri\footrecall{isglobal}\footrecall{cism}\footremember{icl}{Imperial College London: South Kensington Campus, London SW7 2AZ, U.K.}

\end{large}
\end{center}

\vspace{5mm}

\begin{center}
\textbf{Abstract}  
\end{center}

\vspace{5mm}

\begin{center}
\begin{changemargin}{3cm}{3cm}

Foreign direct investment (FDI) in Mozambique has increased rapidly in the last two decades. The growing interest in corporate social responsibility (CSR) - combined with a recent push for malaria elimination - suggest the need for a critical examination of the role of FDI in malaria control and elimination. Through a systematic review of the literature, we find that there has been a notable increase in research pertaining to FDI, CSR and malaria in recent years. Consensus among researchers is that this increase has not been accompanied by a substantial private sector investment in malaria control (with a few notable exceptions). This suggests a potential opportunity for scaling up privately driven malaria control as a means to both improve public health and increase return on investment. However, given the lack of coordination between public and private sectors, even when interests coincide, an over-reliance on foreign and private initiative for funding eradication is not without risks.

\end{changemargin}
\end{center}

@@

Mozambique is a Southern-East African country with a surface area of
nearly 800,000 m2 (more than 3 times that of the United Kingdom) and a
population of nearly 30 million inhabitants. Though low-income,
Mozambique experienced an upward trend in gross domestic product (GDP)
per capita from approximately 200 USD in the late 90s to 623 in 2014,
followed by a steep decrease in 2015 down to 382 in 2016 (WB 2016).

Mozambique's economic growth up to 2014 was facilitated by the
government's open policy towards foreign direct investment (FDI), a
plentiful supply of natural resources (Rogers 2014), and relatively
inexpensive labor. Foreign investment in Mozambique is largely geared
towards the export market, and heavily concentrated. 63\% of exports
come from aluminum, electricity, minerals, and gas, with each sector
displaying high degrees of concentration (greater than 50\% of the
export share of the previous four industries are attributable to one
company) (Sutton 2014). The total number of foreign enterprises
operating in Mozambique is not ascertainable, but in both quantity and
diversity of sources, FDI has increased dramatically in recent years
(Sutton 2014). Mozambique was the Sub-Saharan African country with the
greatest increase in FDI (defined by the World Bank as cross-border
investment to establish a lasting interest) inflows from 2006 to 2014,
registering a boom between 2010 and 2014 of about 5000 million USD
(UNCTAD 2012).

The impact of FDI on economic growth and other economic indicators has
been largely assessed (Alfaro et al. 2006) (Almfraji and Almsafir 2014)
(Brundtland 1999). As FDI has been found to improve working conditions
in low-income countries and, consequently, population ability to pay
(Bloom and Canning 2008), they are likely to improve access to
healthcare and, therefore, improve health (Feenstra and Hanson 1997).
However, there is some evidence that FDI may have adverse effects on
health, through increasing the consumption of ``bads'' such as tobacco
and alcohol and the level of harmful pollution (Pazienza 2015) (Shahbaz
et al. 2015). It has been recently assessed that FDIs are weakly
associated with a marginal benefit in adults' life expectancy in low and
middle income countries (no impact has been found on children's health),
yet investments into the secondary sector (for being polluting) have
been found potentially harmful (Burns et al. 2017).

In Mozambique, improvements in health have accompanied economic
expansion, but the country still lags behind in many basic health
outcomes (Williams et al. 2015). For example, clusters exist in Southern
Mozambique with HIV prevalence as high as 50\% (González et al. 2015).

Malaria is a protozoan parasitic disease, transmitted to humans by
mosquitoes. The \emph{Plasmodium falciparum} species, transmitted by the
female \emph{Anopheles} mosquito, accounts for a large majority of
deaths (N. J. White et al. 2014). Estimates for the year 2015 quantify
the burden of malaria in Mozambique in 8,300,000 cases and 15,000
associated deaths (WHO 2016). Despite the still high malaria burden,
Mozambique is part of the ``Elimination Eight (E8)'' malaria initative,
with the country-specific goal of malarai elimination by 2030.
Elimination efforts have been particularly active in Southern
Mozambique, in border areas contiguous with Swaziland and South Africa,
where the goal is elimination by 2020 (Moonasar et al. 2016).

Several economic activities are strictly linked to malaria. On the one
hand, particularly agricultural-based and extraction activities are
responsible for considerable increases in the malaria burden, through
land and environment modifications that favor the efficiency of
mosquitos in spreading the malaria parasites (Sheela et al. 2017). On
the other hand, the same activities, are barely carried out without
consistent investment towards malaria management, both prevention and
prompt treatment: firms need to protect their workers against infections
to maintain their productivity high. In this regards, there is little
evidence on economic burden malaria places on the private sector in
endemic countries: however, available reports point to profitability of
founding malaria control (Nonvignon et al. 2016). Malaria control
carried out by some multinational companies have been highly successful
and strategies implemented were adopted as a model for the whole country
(Asante et al. 2011).

Private investments in malaria control can be done as part of the
``normal actions'' of private sector or can go slightly beyond what is
strictly needed for having workers' health to carry out activities and
be seen as part of the corporate social responsibility (CSR). CSR is
defined as those cases in which a company goes ``beyond the legal
requirements of the country in which they operate'' in their work
towards that country's long-term interests, in such a systematic manner
that it becomes ``business as usual'' (Ortega et al. 2016).

Motivations for engaging in CSR has been analyzed in the economic
literature. Enterprises would have no incentives to engage in CSR
according to Friedman (Friedman 1970) as firms are organized to maximize
their profits only. The opposite view is that CSR is actually a strategy
for maximizing profits. According to this vision, CSR helps enterprises
to make their products known and, therefore, foment consumers' demand
(Porter and Kramer 2002). CSR also helps building a positive reputation
of enterprises among institutions and communities where they operate:
this is particularly useful for foreign firms that have no previous
contacts with the host country/communities (Zhu, Sun, and Leung 2013).

A study has recently looked at the propensity and at the magnitude of
CSR to the host communities in the US by comparing foreign and local
enterprises (Blonigen and O'Fallon 2011). They found that foreign-owned
enterprises are less likely to give, but that when they do give, it is
substantially more in magnitude than domestic firms, everything else
equal. Their null hypothesis was that foreign firms give as much as
local ones based on the assumptions that firms decide on CSR decisions
from a strict profit maximization perspective and that, therefore,
foreign or local ownership does not affect such a choice. One
alternative hypothesis was that foreign firms are less likely to give or
give less than local because their products are mainly not for the local
markets and because foreign managers have higher propensity to give not
locally but either internationally or to their country of origin. A
second alternative hypothesis is that foreign firms give more than the
local ones. This may be supported by the fact that foreign companies
have different culture from local, more in favor of giving or because
foreign firms use CSR to overcome political and cultural barriers in the
host country that would not be able to mitigate differently.

The relationship between FDI and foreign aid may also be an interesting
determinant of CSR: if aid from a particular country increases FDI from
the same country to the same host communities, incentives to engage in
CSR from foreign private companies may disappear (Selaya and Sunesen
2012). In other words, why should foreign companies donate to the host
country when the government of the country of origin is already giving?

Private foreign investors showed a great optimism during the
International Monetary Fund meeting held in Maputo, the Mozambican city
capital, in 2014 (Pfeiffer et al. 2017). Beginning in late 2015, the
flood of FDI into Mozambique (and many other subsaharan African nations)
slowed to a trickle. The impact of this slowdown on CSR is not yet
known, but it can reasonably be assumed that it will mean a reduction in
CSR activities (albeit with lag). Almost in parallel to this, Mozambique
has had serious financial problems which have resulted in default
(Economist 2017). Given the rapidly changing economic and epidemiologic
context in Mozambique, a comprehensive and current understanding of both
(a) the landscape of FDI and CSR in Mozambique and (b) public health
issues (namely, malaria control) which are directly affected those
investments is sorely needed. Such an understanding may foster greater
private interest in for-profit investment in public health measures.
Likewise, it can facilitate better public sector understanding of
potential industry partners and stakeholders (a prerequisite to greater
public-private collaborations), and guide the public sector away from an
inflexible dependence on FDI for the provision of public health
necessities.

@@

We examined trends from multiple datasets related to FDI, CSR and
malaria in Mozambique. These included the ``Doing Business'' data
pertaining to the measurement of regulatory efficiency (WB 2016), the
GADM (Gadm 2009), the Deutsche Bundesbank Data Repository for foreign
exchange rate history (Bundesbank 2015), the Knoema World Data Atlas for
macroeconomic trends (Knoema 2015), the World Bank data portal for
information containing to net foreign inflows and FDI, USAID Demographic
and Health Survey data on sociodemographics and health-related practices
(WB 2017), the Instituto Nacional de Estatistica for granular data
pertaining to Mozambique's population and economic activities (INE
2017), the Institute for Health Metrics and Evaluation data pertaining
to disease trends over time (IHME 2015), the International Monetary
Fund's open datasets pertaining to FDI (IMF 2003), and the United
Nations Conference on Trade and Development's data on FDI (UNCTAD 2012).

@@

According to the IMF, 69 countries had foreign direct investment in
Mozambique as of the end of
2015\footnote{U.A.E., United States, South Africa, Mauritius, Italy, Australia, Portugal, U.K., India, Netherlands, Japan, Switzerland, Malaysia, France, Germany, Virgin Islands, British, Tanzania, Brazil, Bahamas, The, China, Spain, Norway, Botswana, Kuwait, Guinea-Bissau, Zimbabwe, Togo, China, P.R.: Hong Kong, Puerto Rico, Austria, Sweden, Malta, Isle of Man, Korea, Republic of, Lebanon, Singapore, Angola, China, P.R.: Macao, Denmark, Swaziland, Kenya, Estonia, Suriname, Zambia, Canada, Thailand, Ghana, Turkey, Seychelles, Argentina, Slovak Republic, Serbia, Republic of, Guadeloupe, Reunion, Panama, Ireland, Nigeria, Uruguay, Saudi Arabia, Indonesia, Romania, Bulgaria, Bermuda, Russian Federation, Burundi, Congo, Democratic Republic of, Iceland, Uganda, Cyprus}.
The below chart shows the top 10 countries ranked by amount invested in
Mozambique.

\begin{center}\includegraphics{figures/unnamed-chunk-7-1} \end{center}

What stands out is the presence of many ``non-Western'' countries in the
top 10. Though reliable data only go as far back as 2009, the landscape
of investment 20 years prior would have been both (a) far smaller and
(b) primarily composed of western countries, especially given that
Mozambique emerged so late from civil war (1994). Though as of 2015,
China did not figure in the top 10 FDI countries in Mozambique, its
upward trend is notable, with approximately 1.5 million USD invested in
2010 to nearly 100 million by 2015 (a 60-fold increase).

@@

\begin{center}\includegraphics{figures/unnamed-chunk-10-1} \end{center}

The massive increases in FDI have also been facilitated by dramatic
decreases in the costs to import and export, as well as the costs of
starting a business and registering property (WB 2016).

\begin{center}\includegraphics{figures/unnamed-chunk-11-1} \end{center}

@@

\begin{center}\includegraphics{figures/unnamed-chunk-12-1} \end{center}

@@

\begin{center}\includegraphics{figures/unnamed-chunk-13-1} \end{center}

@@

\begin{center}\includegraphics{figures/unnamed-chunk-14-1} \end{center}

FDI is associated with CSR. Whether under the guise of CSR or not, it is
noteworthy that the private sector currently does play a role in malaria
control activities. According to 2011 DHS data, greater than 7 in every
1,000 households had a private company carry out indoor residual
spraying (DHS 2011). And, in some clusters, the percent of houses
covered by private IRS was greater than 25\%.

@@

\begin{center}\includegraphics{figures/unnamed-chunk-16-1} \end{center}

@@

Firms with CSR activities are often large and foreign. Among the largest
``key players'' in Mozambican CSR are Coca-Cola, British Petroleum and
Colgate-Palmolive. State conglomerates, such as Águas de Moçambique and
Electricidade de Moçambique, also participate in CSR activities (Compact
2007). Large communications firms, such as Vodacom and MCEL, and banks
(BIM and BCI) have a CSR component, but generally focus on culture,
environment and sport, avoiding activities which could either complement
or conflict with the health sector (Kaufmann and Simons-Kaufmann 2015).

Among major foreign companies, those operating in extraction -
particularly mining - appear to be the most involved with malaria. BHP
Billiton has been active in vector control support to the governments of
Mozambique, Swaziland, and South Africa, particularly in the early 2000s
(BHP 2014). Kropfmuhl AG and Vale, also mining companies, have carried
out CSR related to both community development and malaria control in
Northern Mozambique (Kaufmann and Simons-Kaufmann 2015).

Most CSR-funded activities are focused on education, community
development, women's rights and entrepreneurship. According to a UN
poll, none of the country's largest firms invest directly in
malaria-related CSR activities (Compact 2007). That said, a majority of
the firms interviewed by the UN indicated that one of the principal
reasons for investment in CSR is to complement government efforts. To
the extent that malaria accounts for more of the loss in
disability-adjusted life years in Mozambique than comparable countries
(IHME 2015), malaria control's lack of representation among core CSR
activities is a notable absence. One important exception is the
Nando's-lead Goodbye Malaria Trust, an umbrella organization which
inclues a development impact bond, the Goodbye Malaria initiative, and
partnerships with several other foreign firms operating in Mozambique.

@@

\begin{center}\includegraphics{figures/unnamed-chunk-19-1} \end{center}

@@

The intensity of CSR activity seems to closely track FDI, but generally
avoids malaria control. The few malaria-related CSR projects in recent
years has come in large, ``mega-projects'' whose well-publicized malaria
abatement activities are profit-driven (Mouzin and al. 2011) or whose
primary aim is not profit-related (Han 2015). Though the former is often
portrayed as a ``win-win'' for business and public health, the latter
also offers tangible benefits for private industry, and should be
understood as operating under the same conditions and with the same
motivations. To the extent that most CSR activity has avoided the health
sector (including malaria) in favor of cultural, environmental,
education and more general ``development'' activities, it is reasonable
to assume that this may be due to the perceived costs of malaria
control, lack of perceived PR benefits, the government and NGO's
predominant role in the area, as well as the ``opt-in'' nature and
privacy/legal issues generally related to engaging in health-related
campaigns.

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

@@

\hypertarget{ref-Alfaro_2006}{}
Alfaro, Laura, Areendam Chanda, Sebnem Kalemli-Ozcan, and Selin Sayek.
2006. ``How Does Foreign Direct Investment Promote Economic Growth?
Exploring the Effects of Financial Markets on Linkages,'' September.
National Bureau of Economic Research.
doi:\href{https://doi.org/10.3386/w12522}{10.3386/w12522}.

\hypertarget{ref-Almfraji_2014}{}
Almfraji, Mohammad Amin, and Mahmoud Khalid Almsafir. 2014. ``Foreign
Direct Investment and Economic Growth Literature Review from 1994 to
2012.'' \emph{Procedia - Social and Behavioral Sciences} 129 (May).
Elsevier BV: 206--13.
doi:\href{https://doi.org/10.1016/j.sbspro.2014.03.668}{10.1016/j.sbspro.2014.03.668}.

\hypertarget{ref-Asante_2011}{}
Asante, Kwaku P, Charles Zandoh, Dominic B Dery, Charles Brown, George
Adjei, Yaw Antwi-Dadzie, Martin Adjuik, et al. 2011. ``Malaria
Epidemiology in the Ahafo Area of Ghana.'' \emph{Malaria Journal} 10
(1). Springer Nature: 211.
doi:\href{https://doi.org/10.1186/1475-2875-10-211}{10.1186/1475-2875-10-211}.

@@

\hypertarget{ref-bhp}{}
BHP. 2014. ``BHP Billiton Case Study: Malaria.''
\url{https://www.bhp.com/-/media/bhp/documents/investors/reports/2004/norvatispresentation.pdf}.

\hypertarget{ref-NBERw17282}{}
Blonigen, Bruce, and Cheyney O'Fallon. 2011. ``Foreign Firms and Local
Communities.'' Working Paper 17282. Working Paper Series. National
Bureau of Economic Research.
doi:\href{https://doi.org/10.3386/w17282}{10.3386/w17282}.

@@

\hypertarget{ref-Burns_2017}{}
Burns, Darren K., Andrew P. Jones, Yevgeniy Goryakin, and Marc Suhrcke.
2017. ``Is Foreign Direct Investment Good for Health in Low and Middle
Income Countries? An Instrumental Variable Approach.'' \emph{Social
Science \& Medicine} 181 (May). Elsevier BV: 74--82.
doi:\href{https://doi.org/10.1016/j.socscimed.2017.03.054}{10.1016/j.socscimed.2017.03.054}.

@@

\hypertarget{ref-Economist2017}{}
Economist. 2017. ``Mozambique's Default: Mozambique Fails to Pay Its
Debts.'' \emph{The Economist}, January. Economist.
\url{https://www.economist.com/news/middle-east-and-africa/21715030-mozambique-fails-pay-its-debts-mozambiques-default}.

\hypertarget{ref-Feenstra}{}
Feenstra, Robert, and Gordon Hanson. 1997. ``Foreign Direct Investment
and Relative Wages: Evidence from Mexico's Maquiladoras.'' \emph{Journal
of International Economics} 42 (3-4): 371--93.
\url{https://EconPapers.repec.org/RePEc:eee:inecon:v:42:y:1997:i:3-4:p:371-393}.

\hypertarget{ref-Friedman}{}
Friedman, Milton. 1970. ``The Social Responsibility of Business Is to
Increase Its Profits.'' \emph{Corporate Ethics and Corporate
Governance}. Springer Berlin Heidelberg, 173--78.
doi:\href{https://doi.org/10.1007/978-3-540-70818-6_14}{10.1007/978-3-540-70818-6\_14}.

\hypertarget{ref-gadm}{}
Gadm. 2009. ``GADM Database of Global Administrative Areas Version
1.0.''

@@

\hypertarget{ref-Gonz_lez_2015}{}
González, Raquel, Orvalho J. Augusto, Khátia Munguambe, Charlotte
Pierrat, Elpidia N. Pedro, Charfudin Sacoor, Elisa De Lazzari, et al.
2015. ``HIV Incidence and Spatial Clustering in a Rural Area of Southern
Mozambique.'' Edited by Jean KEditor Carr. \emph{PLOS ONE} 10 (7).
Public Library of Science (PLoS): e0132053.
doi:\href{https://doi.org/10.1371/journal.pone.0132053}{10.1371/journal.pone.0132053}.

@@

\hypertarget{ref-imf}{}
IMF. 2003. ``International Monetary Fund's Foreign Direct Investment
Trends and Statistics.''
\url{https://www.imf.org/external/np/sta/fdi/eng/2003/102803.htm}.

@@

\hypertarget{ref-ine}{}
---------. 2017. ``Instituto Nacional de Estatistica.''
\url{http://www.ine.gov.mz/}.

\hypertarget{ref-Kaufmann_2015}{}
Kaufmann, Friedrich, and Claudia Simons-Kaufmann. 2015. ``Corporate
Social Responsibility in Mozambique.'' \emph{CSR, Sustainability, Ethics
\& Governance}, December. Springer International Publishing, 31--50.
doi:\href{https://doi.org/10.1007/978-3-319-26668-8_2}{10.1007/978-3-319-26668-8\_2}.

@@

\hypertarget{ref-Moonasar_2016}{}
Moonasar, Devanand, Rajendra Maharaj, Simon Kunene, Baltazar Candrinho,
Francisco Saute, Nyasatu Ntshalintshali, and Natashia Morris. 2016.
``Towards Malaria Elimination in the Mosaswa (Mozambique, South Africa
and Swaziland) Region.'' \emph{Malaria Journal} 15 (1). Springer Nature.
doi:\href{https://doi.org/10.1186/s12936-016-1470-8}{10.1186/s12936-016-1470-8}.
